---
title: "Biogeography_Project"
author: "Devin Richards"
date: "2024-12-02"
output: html_document
---
### Required packages & Directory:
```{r, warning=FALSE, message=FALSE}
library(sf)
library(ggplot2)
library(dplyr)
library(tidyr)
library(rmapshaper)
library(rnaturalearth)
library(rnaturalearthdata)
library(classInt)
library(gridExtra)
library(eHOF)
setwd("D:\\Biogeography\\Data Files")
```
### Sourced Data Files:
```{r, results='hide'}
ecoregions <- st_read("D:\\Biogeography\\Data Files\\na_cec_eco_l2")
iv_data <- readRDS("D:\\Biogeography\\Data Files\\df_master.rds")
```
### First step is to isolate the Appalachia ecoregion from all of North America
```{r}
head(ecoregions)
head(unique(ecoregions$NA_L2NAME))
appalachia <- ecoregions %>%
  filter(NA_L2NAME == "OZARK/OUACHITA-APPALACHIAN FORESTS")
head(appalachia)
```
### There is a problem with the ecoregion file, the regions are polygons & projected via Lambert Azimuthal Equal Area. We need to project them as WGS 1984, and extract Lat + Lon data.
```{r}
# Creating Lat & Lon coordinates for the Appalachia only:
appalachia_wgs84 <- st_transform(appalachia, crs = 4326)
appalachia_cords <- st_coordinates(appalachia_wgs84)
app_cords_df <- as.data.frame(appalachia_cords)
colnames(app_cords_df) <- c("X", "Y")
head(app_cords_df)
```
### Now I can plot the Appalachia region over the base map of our 31 states in the FIA database.
```{r}
# Adding eastern US states shape file for background
usa <- ne_states(country = "United States of America", returnclass = "sf")
usa31 <- usa %>%
  filter(longitude >= -95 & longitude <= -65)

ggplot() +
  geom_sf(data = usa31, fill = "lightblue", color = "black") +
  geom_point(data = app_cords_df, aes(x = X, y = Y, color = "red"), size = 0.5)
```
### Next, I want to create a function to plot a specific tree species over its latitudinal extent and symbolizing it with points. The symbolized points will be normalized jenks of that trees IV.
### This is what I am currently working on: Below is the code from class that results in a graph of IV over latitude.
```{r, results='hide'}
red_maple_data <- iv_data %>%
  filter(common_name == "red maple")
head(red_maple_data)
min(red_maple_data$LAT)
max(red_maple_data$LAT)

# Create new empty data frame, to average IV to each 1 degree latitudinal band
lat_seq <- seq(from = floor(min(red_maple_data$LAT)), to = ceiling(max(red_maple_data$LAT)), by = 0.5)
lat_seq
red_maple_bands <- data.frame(
  Lat_band = lat_seq,
  avg_IV = numeric(length(lat_seq)))
head(red_maple_bands)

test <- red_maple_data %>%
  filter(LAT >= 30 & LAT <= 35) %>%
  mutate(avg_IV = mean(IV, na.rm = TRUE))
head(test)

lat_seq
for (i in 1:length(lat_seq)) {
  lat_min <- lat_seq[i] - 0.5
  lat_max <- lat_seq[i] + 0.5
  band_data <- red_maple_data %>%
    filter(LAT >= lat_min & LAT <= lat_max)
  red_maple_bands$avg_IV[i] <- mean(band_data$IV, na.rm = TRUE)
}
head(red_maple_bands)
tail(red_maple_bands)
red_maple_bands <- red_maple_bands[-1,]

hof_model <- HOF(
  red_maple_bands$avg_IV,
  red_maple_bands$Lat_band,
  modeltypes = c("I","II","III","IV","V"),
  family = gaussian,
  bootstrap = 100,
  test = 'AIC'
)
hof_model
best_model <- pick.model(hof_model, modeltypes = c("I","II","III","IV","V"), test = 'AIC')
best_model
predicted_response <- predict(hof_model, model = best_model,
                              newdata = seq(min(red_maple_bands$Lat_band), max(red_maple_bands$Lat_band), by = 0.1))
predicted_response

seq(min(red_maple_bands$Lat_band), max(red_maple_bands$Lat_band), by = 0.1)

scaled_response <- predicted_response * max(red_maple_bands$avg_IV)
scaled_response
```
```{r}
ggplot(red_maple_bands, aes(x = Lat_band, y = avg_IV)) +
  geom_point(color = "black") +
  geom_line(data = data.frame(lat_band = seq(min(red_maple_bands$Lat_band), max(red_maple_bands$Lat_band), by = 0.1),
                              predicted_IV = scaled_response),
            aes(x = lat_band, y = predicted_IV), color = "purple") +
  geom_vline(xintercept = Para(hof_model, model = best_model)$opt, color = 'red', linetype = "dashed") +
  labs(title = "Latitudinal IV pattern for Red Maple", x = "Latitude", y = "IV (Importance Value)")

Para(hof_model, model = best_model)$opt
```